---
title: "Recoding Data"
output: html_document
---

```{r echo=FALSE}
library(printr)
```

```{r, message = FALSE}
library(readr)
library(dplyr)

```

Suppose our dataset looks something like this

```{r}
fruits <- read_csv(
"name, quantity
gala apples, 3
red delicious, 2
bananas, 4
oranges, 6
grape-fruits, 2
blueberries, 20
straw-berries, 15
strawberries, 12
")

fruits
```

Let's assume that we  wanted to correct the spelling or standardize the value of an existing column. For example, in our case, `strawberries` are sometimes spelled with a hyphen and sometimes without a hyphen. And sometimes the dataset uses specific variety of apples instead of just using `apples` so we need to fix that as well.

### Solution Without `dplyr`

First let's make a copy of the `name` variable to ensure that we don't overwrite the original variable by mistake.

```{r}
fruits$new_name <- fruits$name

fruits
```

Now let's fix apples:

```{r}
fruits$new_name[fruits$name == "gala apples" | fruits$name == "red delicious"] = "apples"

fruits
```

You can repeat this as many time as you want, and if you have more than one value to compare against then it's best to use `%in%`

```{r}
fruits$new_name[fruits$name %in% c("grapefruits", "grape-fruits")] = "grapefruits"
fruits$new_name[fruits$name %in% c("straw-berries", "strawberries")] = "strawberries"

fruits
```

### The `dplyr` Solution

Here's how you'd do the same with `dplyr` using `mutate` and `ifelse`. First let's recreate the dataset

```{r}
fruits <- read_csv(
"name, quantity
gala apples, 3
red delicious, 2
bananas, 4
oranges, 6
grape-fruits, 2
blueberries, 20
straw-berries, 15
strawberries, 12
")

fruits
```

Then we copy `name` to `new_name`, and call `mutate()` for as many values as we need to recode.

```{r}
fruits <- fruits %>%
  mutate(new_name = name) %>%
  mutate(new_name = ifelse(name %in% c("gala apples", "red delicious"), "apples", new_name)) %>%
  mutate(new_name = ifelse(name %in% c("grapefruits", "grape-fruits"), "grapefruits", new_name)) %>%
  mutate(new_name = ifelse(name %in% c("straw-berries", "strawberries"), "strawberries", new_name))

fruits
```

As you can see, both the solutions give us exactly the same results, so you can use whichever one you prefer. 

