---
title: "Data Reshape Example"
output: html_document
---

```{r echo=FALSE}
library(printr)
```

```{r message=FALSE}
library(dplyr)
library(tidyr)
```

Create a small dataset to test the reshape code.

```{r}
events <- data_frame(
  event_id = c(1001, 1002, 1003),
  side_a = c("blue", "red", "green"),
  side_b = c("red", "green", "blue"),
  outcome = c("victory", "victory", "peace"),
  winner = c(1, 2, 1)
)
```

The data we start with has three events:

```{r}
events
```

We reshape based on who won. The steps are:

1. Create two new binary variables `side_a_won` and `side_b_won` that indicate whether each side won
1. Gather the newly created variables into a a key-value pair of `side` and `win_status` variables
1. Modify the `side` variable to include the name of the actor
1. Create two separate variables `victory` and `peace` from the `outcome` variable
1. Optional: Select only the columns we care about
1. Optional: Use `arrange()` to make sure the resulting dataframe maintains the same order as the initial

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2) %>%
  gather(side, win_status, side_a_won, side_b_won)  %>%
  mutate(side = ifelse(side == "side_a_won", side_a, side_b)) %>%
  mutate(victory = outcome == "victory", 
         peace = outcome == "peace") %>%
  select(event_id, side, win_status, victory, peace) %>%
  arrange(event_id)
```  

And finally we get two observations for each event but a single actor per observation with their victory and outcome status.

Let's walk through each step one at a time:

1\. Create two new binary variables `side_a_won` and `side_b_won` that indicate whether each side won

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2)
```  

2\. Gather the newly created variables into a a key-value pair of `side` and `win_status` variables

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2) %>%
  gather(side, win_status, side_a_won, side_b_won)
```  

3\. Modify the `side` variable to include the name of the actor

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2) %>%
  gather(side, win_status, side_a_won, side_b_won)  %>%
  mutate(side = ifelse(side == "side_a_won", side_a, side_b))
```  

4\. Create two separate variables `victory` and `peace` from the `outcome` variable

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2) %>%
  gather(side, win_status, side_a_won, side_b_won)  %>%
  mutate(side = ifelse(side == "side_a_won", side_a, side_b)) %>%
  mutate(victory = outcome == "victory", 
         peace = outcome == "peace") 
```  

Steps 5 and 6 are pretty simple and don't necessarily need much explanation

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2) %>%
  gather(side, win_status, side_a_won, side_b_won)  %>%
  mutate(side = ifelse(side == "side_a_won", side_a, side_b)) %>%
  mutate(victory = outcome == "victory", 
         peace = outcome == "peace") %>%
  select(event_id, side, win_status, victory, peace) %>%
  arrange(event_id)
```  

A slightly different (and perhaps more flexible) approach would be to use `spread()` for reshaping the `outcome` column.

```{r}
events %>%
  mutate(side_a_won = winner == 1, 
         side_b_won = winner == 2,
         outcome_status = TRUE) %>%
  gather(side, win_status, ends_with("_won"))  %>%
  mutate(side = ifelse(side == "side_a_won", side_a, side_b)) %>%
  spread(outcome, outcome_status, fill = FALSE) %>%
  select(-c(side_a, side_b, winner)) %>%
  arrange(event_id)
```

